\bf Longitudinal Principal Manifold Estimation

### 2024-07-26 jasa/main.tex
add jasa submission directory

\documentclass[12pt]{article}
\usepackage{algorithm2e}
\usepackage{amscd}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{bm}
\usepackage{caption}
\usepackage{color}
%\usepackage{csquotes}
\usepackage{float}
\usepackage[perpage, symbol*]{footmisc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{url}

@@

@@

\bigskip
\begin{abstract}
Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation. %\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
\end{abstract}

@@

%Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}. Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}. For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes. Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Additionally, there are a large number of manual segmentation protocols, each producing distinct results and incurring varying levels of inter-rater and intra-rater variability \citep{boccardiSurveyProtocolsManual2011}. Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}. While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

@@

%\subsection{A Motivating Question from Medical Imaging}

%Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. %\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models. %In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging \citep{scahill2003longitudinal}. %To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice. A thorough description of the data and computation of hippocampus volumes and further examples are given in Section \ref{s:application} and Figure \ref{fig:lhipp_volume_comparison}.

@@

%\subsection{Background and Our Contributions}

%Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing \citep{shen2022robust}, single-cell biology \citep{ding2023learning}, and robotics \citep{gao2024bi}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures. Nonlinear manifold learning methods have been extensively studied in the literature. However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies. These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension. %We use the oldest linear manifold learning method, principal component analysis \citep[PCA,][]{Pearson1901on}, as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.

@@

%In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time. Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}). Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between time points. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

@@

%Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, including time in the dimension reduction process. Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an explicit functional representation of the manifold estimate. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

%Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis (PCA) method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

%Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

@@

%The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection and problems with applications in high-dimensional settings. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure \ref{fig:adni_result}). For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows.

@@

%The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(r_1,r_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

%Hereafter, we utilize the following notations to describe the LPME framework for generic low and high dimensions, $d$ and $D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each individual $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. In this setting, the manifold to estimate is represented by $(t,\mathbf{r})\mapsto F(t,\mathbf{r})\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.

@@

%At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping. Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space. For example, the PME algorithm can be used to obtain this parameterization. In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, smoothing over the spatial spline coefficients with respect to time. This yields a multi-stage model that, for a provided time point and set of tuning parameters, obtains a set of spline coefficients that are time-dependent, then returns the high-dimensional output of the spline function associated with the tuning parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. A visualization of these steps is shown in Figure \ref{fig:lpme_steps}. Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}.

@@

To obtain consistent parameterizations of the obtained cluster centers $\left\{\mu_{j, t}\right\}_{j=1}^{N_t}$ over time $t$, we propose using the PME algorithm to develop a parameterization of the centers $\left\{\mu_{j, t_1}\right\}_{j=1}^{N_{t_1}}$ taken from the first time point $t_1$, then using the projection index associated with this estimate to find initial parameterizations of the centers for the remaining time points. This approach is based on the assumption that the underlying manifolds of interest remains relatively stable over time. %The results given in Section \ref{s:simulations} indicate that this process ensures that the initial parameterization remains consistent over time under the assumption of relative stability of the shape of the structure.

@@

%Depending on the locations of spline knots  across time points, this function may take the form of a case in the varying-coefficient model framework discussed in \cite{hastieVaryingCoefficientModels1993}. However, instead of attempting to estimate each coefficient function individually, we consider all coefficients together as the output of a single smoothing spline function. The initial use of smoothing splines to estimate principal manifolds complicates the direct use of estimation approaches developed for this model class in our use case.

@@

\begin{figure}
  \centering
  %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_plots/adni_lhipp_cross_section}} \label{fig:lhipp_cross_sections}}%
  \hfill
  \subfloat[\centering Left Thalamus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_plots/adni_lthal_cross_section}} \label{fig:lthal_cross_sections}}

  \caption{{\footnotesize Left Hippocampus and Left Thalamus Cross Sections. LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the observed hippocampus surface. In the fifth hippocampus observation with $t=6.55$, the LPME-estimated surface has a gap. Such gaps in the surface estimates frequently correspond with inaccurate volume estimates. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}}
  \label{fig:adni_cross_sections}
\end{figure}

@@

%\begin{figure}[h]
%\centering
%\subfloat{
%\label{fig:adni_lhipp_result}
%\begin{subfigure}
%\includegraphics{adni_plots/adni_lhipp_data_plot}
%\caption{Data}
%\label{fig:adni_lhipp_result_data}
%\end{subfigure}
%}

% \hfill
%\subfloat{
%\label{fig:adni_lthal_result}
%\begin{subfigure}
%\includegraphics{adni_plots/adni_lthal_data_plot}
%\caption{Data}
%\label{fig:adni_lthal_result_data}
%\end{subfigure}
%}

%\caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant. The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue. The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
%\label{fig:adni_result}

%\end{figure}

@@

%Similar to other applications of statistical smoothing techniques, it is difficult to compare performance of models in real data experiments. Specifically, the MSD provides a summary of the distance of the data points and their projections onto the estimated manifolds for each model. This means that a model that performs well using this metric may not actually be the best fit for the underlying manifold, e.g. in the case of overfitting. This largely explains why PME does not usually show better performance than the PC/PS methods using this metric - both approaches are estimating a curve or surface using the data available at a given time point, which may be affected by systematic noise that increases the distance to the true manifold. When comparing the fit of the PME and principal surface and principal curve methods in the simulated data, PME tends to show improvement in the median MSD, with elevated mean MSD levels compared to principal curves.

@@

%\nocite{*}
%\bibliographystyle{plain}
%\bibliography{references}
%\printbibliography

@@

%Data collection and sharing for the Alzheimer's Disease Neuroimaging Initiative (ADNI) is funded by the National Institute on Aging (National Institutes of Health Grant U19 AG024904). The grantee organization is the Northern California Institute for Research and Education. In the past, ADNI has also received funding from the National Institute of Biomedical Imaging and Bioengineering, the Canadian Institutes of Health Research, and private sector contributions through the Foundation for the National Institutes of Health (FNIH) including generous contributions from the following: AbbVie, Alzheimer’s Association; Alzheimer’s Drug Discovery Foundation; Araclon Biotech; BioClinica, Inc.; Biogen; Bristol-Myers Squibb Company; CereSpir, Inc.; Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli Lilly and Company; EuroImmun; F. Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research & Development, LLC.; Johnson & Johnson Pharmaceutical Research &Development LLC.; Lumosity; Lundbeck; Merck & Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics.
%} \fi

>>>>

\documentclass[12pt]{article}
\usepackage{algorithm2e}
\usepackage{amscd}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{bm}
\usepackage{caption}
\usepackage{color}
\usepackage{float}
\usepackage[perpage, symbol*]{footmisc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{url}

@@

\pdfminorversion=4

@@

\bigskip
\begin{abstract}
Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation.
\end{abstract}

@@

@@

@@

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures. Nonlinear manifold learning methods have been extensively studied in the literature. However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies. These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.

@@

@@

@@

@@

@@

@@

To obtain consistent parameterizations of the obtained cluster centers $\left\{\mu_{j, t}\right\}_{j=1}^{N_t}$ over time $t$, we propose using the PME algorithm to develop a parameterization of the centers $\left\{\mu_{j, t_1}\right\}_{j=1}^{N_{t_1}}$ taken from the first time point $t_1$, then using the projection index associated with this estimate to find initial parameterizations of the centers for the remaining time points. This approach is based on the assumption that the underlying manifolds of interest remains relatively stable over time.

@@

@@

\begin{figure}
  \centering
  \subfloat[\centering Left Hippocampus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_lhipp_cross_section}} \label{fig:lhipp_cross_sections}}%
  \hfill
  \subfloat[\centering Left Thalamus Cross Section]{{\includegraphics[width=0.49\textwidth]{adni_lthal_cross_section}} \label{fig:lthal_cross_sections}}

  \caption{{\footnotesize Left Hippocampus and Left Thalamus Cross Sections. LPME-approximated embedding function routinely falls inside the boundaries of the observed hippocampus surface. In the fifth hippocampus observation with $t=2.11$, the LPME-estimated surface has a gap. Such gaps in the surface estimates frequently correspond with inaccurate volume estimates. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}}
  \label{fig:adni_cross_sections}
\end{figure}

@@

@@

@@

@@

%Data collection and sharing for the Alzheimer's Disease Neuroimaging Initiative (ADNI) is funded by the National Institute on Aging (National Institutes of Health Grant U19 AG024904). The grantee organization is the Northern California Institute for Research and Education. In the past, ADNI has also received funding from the National Institute of Biomedical Imaging and Bioengineering, the Canadian Institutes of Health Research, and private sector contributions through the Foundation for the National Institutes of Health (FNIH) including generous contributions from the following: AbbVie, Alzheimer’s Association; Alzheimer’s Drug Discovery Foundation; Araclon Biotech; BioClinica, Inc.; Biogen; Bristol-Myers Squibb Company; CereSpir, Inc.; Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli Lilly and Company; EuroImmun; F. Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research \& Development, LLC.; Johnson \& Johnson Pharmaceutical Research \& Development LLC.; Lumosity; Lundbeck; Merck \& Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics.
%} \fi
